---
title: "PlasticDaphnia: Nicola's exposure PCoA ordinations: Daphnia's microbiome"
output: html_notebook
---
```{r}
library(phyloseq)
library(vegan)
library(ggplot2)
library(ggpubr)
library(data.table)
library(BiocManager)
library(microbiome)
library(knitr)
#install.packages("patchwork")
library(patchwork)
```

Load ps-object and metadata
```{r}
#Load phyloseq (rarefied)
ps_mic = readRDS("/Users/u0145079/Library/CloudStorage/OneDrive-KULeuven/Desktop/PlasticDaphnia/Exposure-Niki/tables/phyloseq/decontam/ps_rare_daphnia_decontam.rds")

#load meta
meta = read.csv("/Users/u0145079/Library/CloudStorage/OneDrive-KULeuven/Desktop/PlasticDaphnia/Exposure-Niki/tables/decontaminated_set/merged_meta.csv",row.names = 1)
```

Load metadata into ps
```{r}
# Extract common sample IDs
common_sample_ids <- intersect(sample_names(ps_mic), row.names(meta))
# Subset metadata to include only common sample IDs
filtered_meta <- meta[common_sample_ids, ]
# Assign filtered metadata as sample_data for ps_bac
sample_data(ps_mic) <- filtered_meta
ps_mic <- subset_samples(ps_mic, Exposure_time !="Pre-Env") #remove 2 extra env-samples
```


Is microbial composition different (both ponds) between MPs exposed and exposure control? Daphnia in general

```{r}
subset_phyloseq <- subset_samples(ps_mic, Exposure_time =="Post" &  Plastic != "C" & Age == "Adult" | Exposure_time =="Post" & Plastic == "C" & Age == "Adult")
filtered_meta <- data.frame(sample_data(subset_phyloseq))
bray <- vegdist(as.data.frame(t(otu_table(subset_phyloseq)), method="bray")) #creating dissimilarity matrix with vegdist using bray curtis

permanova_result <- adonis2(bray~ Plastic_presence, data = filtered_meta, na.action = na.omit) #0.39
permanova_result
#Dispersion
dispersion_test <- betadisper(bray, filtered_meta$Plastic_presence,bias.adjust =TRUE)
# Perform permutation test
permutest(dispersion_test, permutations = 999)#0.163
```
Are pre-exposure controls the same as exposure controls (jars without plastic added)?
	PERMANOVA between Exposure control and pre-exposure control
If significant, we leave it, if not -> Merge then into "CONTROL"

```{r}
subset_phyloseq <- subset_samples(ps_mic, Exposure_time =="Pre" | Exposure_time =="Post" & Plastic == "C" & Age == "Adult")
filtered_meta <- data.frame(sample_data(subset_phyloseq))
bray <- vegdist(as.data.frame(t(otu_table(subset_phyloseq)), method="bray")) #creating dissimilarity matrix with vegdist using bray curtis

permanova_result <- adonis2(bray~ Exposure_time, data = filtered_meta, na.action = na.omit) #0.004 **

#Dispersion
dispersion_test <- betadisper(bray, filtered_meta$Exposure_time,bias.adjust =TRUE)
# Perform permutation test
permutest(dispersion_test, permutations = 999)#0.179

```
Which community is richer?

```{r}
tab <-microbiome::alpha(subset_phyloseq, index = "all")
meta <- data.frame(sample_data(filtered_meta)) #Accessing my sample information from the ps object containing rarefied data
tab$Exposure_time = meta$Exposure_time

###Chao1
p.chao1 <- boxplot_alpha(subset_phyloseq, 
                           index = "Chao1",
                           x_var = "Exposure_time",
                           fill.colors = c("Pre" = "blue", 
                                           "Post" = "red"))


p.chao1 <- p.chao1 + theme_pubr() + 
  labs(x="", y="Chao1") +
  theme(axis.text = element_text(size=12),
        axis.title = element_text(size=16),
        legend.position = "none") # This line removes the legend

p.chao1
print(t.test(chao1~Exposure_time, data=tab)) # p-value = 2.633e-15


#####Shannon

p.shannon <- boxplot_alpha(subset_phyloseq, 
                           index = "diversity_shannon",
                           x_var = "Exposure_time",
                           fill.colors = c("Pre" = "blue", 
                                           "Post" = "red"))


p.shannon <- p.shannon + theme_pubr() + 
  labs(x="", y="Shannnon") +
  theme(axis.text = element_text(size=12),
        axis.title = element_text(size=16),
        legend.position = "none") # This line removes the legend

p.shannon
print(t.test(diversity_shannon~Exposure_time, data=tab)) # p-value = 1.184e-11


#####Pielou

p.Pielou <- boxplot_alpha(subset_phyloseq, 
                           index = "evenness_pielou",
                           x_var = "Exposure_time",
                           fill.colors = c("Pre" = "blue", 
                                           "Post" = "red"))


p.Pielou <- p.Pielou + theme_pubr() + 
  labs(x="", y="Pielou") +
  theme(axis.text = element_text(size=12),
        axis.title = element_text(size=16),
        legend.position = "none") # This line removes the legend

p.Pielou
print(t.test(evenness_pielou~Exposure_time, data=tab)) # p-value = 0.3529

#####Combine in one plot
p_combined <- p.chao1 / p.shannon / p.Pielou
p_combined
# Optionally, if you want to add a common title or adjust spacing, you can do:
p_combined + 
  plot_annotation(title = "Diversity Indices Comparison for Microbiome samples") +
  plot_layout(guides = 'collect') # This collects and positions the legends together if applicable
```

Yes there is a statistically significant difference between pre-exposure indv. and those not exposed at all to plastic.
Likely due to the uptake of bacterioplankton from the surrounding water. 
The clone controls microbiome changed over the course of exposure in the absence of MPs. There is no difference in the richness.

Is there a difference between BP and DG post-exposure communities? CONTROLS - Non-plastic exposed samples

```{r}
subset_phyloseq <- subset_samples(ps_mic, Pond == "BP" & Category != "Daphnia: Post-Long-exposure (selected survivals)" & Plastic_presence == "No" | Pond == "DG" & Category != "Daphnia: Post-Long-exposure (selected survivals)" & Plastic_presence == "No")
filtered_meta <- data.frame(sample_data(subset_phyloseq))
bray <- vegdist(as.data.frame(t(otu_table(subset_phyloseq)), method="bray")) #creating dissimilarity matrix with vegdist using bray curtis

permanova_result <- adonis2(bray~ Pond, data = filtered_meta, na.action = na.omit) #0.001 ***
permanova_result
#Dispersion
dispersion_test <- betadisper(bray, filtered_meta$Pond,bias.adjust =TRUE)
# Perform permutation test
permutest(dispersion_test, permutations = 999)#0.834

```
```
ANSWER: Yes, there is a difference between DG and BP communities in the absence of MPs exposure
```

Is there a difference between BP and DG post-exposure communities? WITH Plastics

```{r}
subset_phyloseq <- subset_samples(ps_mic, Pond == "BP" & Category != "Daphnia: Post-Long-exposure (selected survivals)" & Plastic_presence == "Yes" | Pond == "DG" & Category != "Daphnia: Post-Long-exposure (selected survivals)" & Plastic_presence == "Yes")
filtered_meta <- data.frame(sample_data(subset_phyloseq))
bray <- vegdist(as.data.frame(t(otu_table(subset_phyloseq)), method="bray")) #creating dissimilarity matrix with vegdist using bray curtis

permanova_result <- adonis2(bray~ Pond, data = filtered_meta, na.action = na.omit) #0.001 ***
permanova_result
#Dispersion
dispersion_test <- betadisper(bray, filtered_meta$Pond,bias.adjust =TRUE)
# Perform permutation test
permutest(dispersion_test, permutations = 999)#0.206

```
```
ANSWER: Yes, there is a difference between DG and BP communities in the presence of MPs
```

BLAUWE POORT:

		○ Only adults and not-long exposure jars
		○ Post exposure only
    ○ Before exposure vs After exposure

```{r}
subset_phyloseq <- subset_samples(ps_mic, Pond == "BP" & Age == "Adult" & Category != "Daphnia: Post-Long-exposure (selected survivals)"| Exposure_time == "Pre")
filtered_meta <- data.frame(sample_data(subset_phyloseq))
bray <- vegdist(as.data.frame(t(otu_table(subset_phyloseq)), method="bray")) #creating dissimilarity matrix with vegdist using bray curtis

permanova_result <- adonis2(bray~ Plastic_presence, data = filtered_meta, na.action = na.omit) #0.016 *
permanova_result
#Dispersion
dispersion_test <- betadisper(bray, filtered_meta$Plastic_presence,bias.adjust =TRUE)
# Perform permutation test
permutest(dispersion_test, permutations = 999)#0.065 .

```

BLAUWE POORT:

		○ Only adults
		○ Post-exposure only
		○ Control(NO MPs in the jar) vs Plastics
		○ If not significant then test for each plastic separetly


```{r}
subset_phyloseq <- subset_samples(ps_mic, Pond == "BP" & Age == "Adult" & Category != "Daphnia: Post-Long-exposure (selected survivals)"| Exposure_time == "Pre")
filtered_meta <- data.frame(sample_data(subset_phyloseq))
bray <- vegdist(as.data.frame(t(otu_table(subset_phyloseq)), method="bray")) #creating dissimilarity matrix with vegdist using bray curtis

permanova_result <- adonis2(bray~ Plastic_presence, data = filtered_meta, na.action = na.omit) #0.016 *
permanova_result
#Dispersion
dispersion_test <- betadisper(bray, filtered_meta$Plastic_presence,bias.adjust =TRUE)
# Perform permutation test
permutest(dispersion_test, permutations = 999)#0.065 .

```


DE GAVERS:

		○ Only adults and not-long exposure jars
		○ Post exposure only
    ○ Plastic present vs Plastic absent

```{r}
subset_phyloseq <- subset_samples(ps_mic, Pond == "DG" & Age == "Adult" & Category != "Daphnia: Post-Long-exposure (selected survivals)"| Exposure_time == "Pre")
filtered_meta <- data.frame(sample_data(subset_phyloseq))
bray <- vegdist(as.data.frame(t(otu_table(subset_phyloseq)), method="bray")) #creating dissimilarity matrix with vegdist using bray curtis

permanova_result <- adonis2(bray~ Plastic_presence, data = filtered_meta, na.action = na.omit) #0.002 **
permanova_result
#Dispersion
dispersion_test <- betadisper(bray, filtered_meta$Plastic_presence,bias.adjust =TRUE)
# Perform permutation test
permutest(dispersion_test, permutations = 999)#0.72 .

```

What are significant effects? All samples

```{r}
subset_phyloseq <- subset_samples(ps_mic, Exposure_time !="Pre" & Plastic != "C")
filtered_meta <- data.frame(sample_data(subset_phyloseq))
bray <- vegdist(as.data.frame(t(otu_table(subset_phyloseq)), method="bray")) #creating dissimilarity matrix with vegdist using bray curtis

permanova_result <- adonis2(bray~ Pond*Clone*Plastic*Age, data = filtered_meta, na.action = na.omit)
permanova_result_df = data.frame(permanova_result)
```

Check for equality of variances:

```{r}
"Pond"
dispersion_test <- betadisper(bray, filtered_meta$Pond,bias.adjust =TRUE)
# Perform permutation test
permutest(dispersion_test, permutations = 999)#0.247

"Age"
dispersion_test <- betadisper(bray, filtered_meta$Age,bias.adjust =TRUE)
# Perform permutation test
permutest(dispersion_test, permutations = 999)#0.007 **
```
Plot PCoA showing diffrences between De-Gavers and Blauwe Poort

```{r}
#NMDS = Exposure + Pond
subset_phyloseq <- subset_samples(ps_mic, Exposure_time != "Pre" & Category != "Daphnia: Post-Long-exposure (selected survivals)")
meta_data <- data.frame(sample_data(subset_phyloseq))
bray <- vegdist(data.frame(t(otu_table(subset_phyloseq))), method="bray") 

PCoA_result <- capscale(bray ~ 1)

# Calculate total variance to normalize eigenvalues
total_variance <- sum(PCoA_result$CA$eig)

# Calculate percentage of variance explained by the first two axes
percent_var1 <- PCoA_result$CA$eig[1] / total_variance * 100
percent_var2 <- PCoA_result$CA$eig[2] / total_variance * 100

meta = data.frame(sample_data(subset_phyloseq))

p <- plot_ordination(subset_phyloseq, PCoA_result, color = "Pond", shape = "Plastic") +  
  geom_point(size = 3) +  
  scale_shape_manual(values = c(4, 16, 17, 15)) +  # Distinguishable shapes, 19 for NA
  scale_colour_manual(values = c("red", "blue"), na.value = "grey50") +  
  labs(title = "", 
       x = paste0("PCoA1 (", format(percent_var1, digits = 2), "%)"), 
       y = paste0("PCoA2 (", format(percent_var2, digits = 2), "%)"), 
       color = "Bacterioplankton inoculum",
       shape = "Control") +  # Change MP treatment to Control
  theme_pubr()


# Perform envfit to fit taxa onto the ordination
taxa_scores <- scores(PCoA_result, display = "species")
envfit_result <- envfit(PCoA_result ~ ., data = as.data.frame(t(taxa_scores)), permutations = 999)

# Extract envfit vectors
scores_envfit <- as.data.frame(scores(envfit_result, display = "vectors"))

# Add envfit vectors
p + geom_segment(data = scores_envfit, 
                 aes(x = 0, y = 0, xend = PCoA1, yend = PCoA2), 
                 arrow = arrow(length = unit(0.3, "cm")), color = "grey") +
  geom_text(data = scores_envfit, 
            aes(x = PCoA1, y = PCoA2, label = rownames(scores_envfit)), 
            color = "black", vjust = 1.5)


```

```{r}
#load seqPCoA.R and Silhouette.R from seqgroup
subset_phyloseq_glom=tax_glom(subset_phyloseq,"Genus")
abundance=data.frame(otu_table(subset_phyloseq))
meta_data <- data.frame(sample_data(subset_phyloseq))
groups=as.vector(meta_data$Plastic)
taxa=data.frame(tax_table(subset_phyloseq))
all(rownames(taxa) == rownames(abundance))

# Filter the abundance matrix to match the filtered taxa
filtered_abundance <- abundance[!(is.na(taxa$Genus) & is.na(taxa$Species)), ]

# Merge Genus and Species and make them unique row names
rownames(filtered_abundance) <- make.unique(paste(filtered_taxa$Genus, filtered_taxa$Species, sep = "_"))

seqPCoA(filtered_abundance,groups=meta_data$Plastic,topTaxa=100,arrowFactor=0.002)

```

What are significant effects? Blauwe Poort
```{r}
subset_phyloseq <- subset_samples(ps_mic, Exposure_time !="Pre" & Plastic != "C" & Pond == "BP")
filtered_meta <- data.frame(sample_data(subset_phyloseq))
bray <- vegdist(as.data.frame(t(otu_table(subset_phyloseq)), method="bray")) #creating dissimilarity matrix with vegdist using bray curtis

permanova_result <- adonis2(bray~ Clone*Plastic*Age, data = filtered_meta, na.action = na.omit)
permanova_result_df = data.frame(permanova_result)
```

```{r}
"Age"
dispersion_test <- betadisper(bray, filtered_meta$Age,bias.adjust =TRUE)
# Perform permutation test
permutest(dispersion_test, permutations = 999)#0.059 **
```
What are significant effects? De Gavers
```{r}
subset_phyloseq <- subset_samples(ps_mic, Exposure_time !="Pre" & Plastic != "C" & Pond == "DG")
filtered_meta <- data.frame(sample_data(subset_phyloseq))
bray <- vegdist(as.data.frame(t(otu_table(subset_phyloseq)), method="bray")) #creating dissimilarity matrix with vegdist using bray curtis

permanova_result <- adonis2(bray~ Clone*Plastic*Age, data = filtered_meta, na.action = na.omit)
permanova_result_df = data.frame(permanova_result)
```


